`SCPCL000824` - Exploration of gene set based methods for tumor cell assignment

### 2024-07-16 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
update output tables for 822 and 824

# output files 
final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000490")
fs::dir_create(final_annotations_dir)
final_annotations_file <- file.path(final_annotations_dir, "SCPCL000822_tumor-classifications.tsv")
```

@@

# generate marker genes df
plot_markers_df <- create_marker_gene_df(
  classification_df, 
  marker_genes_file
)
```

@@

ggplot(marker_density_df, aes(x = sum_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method))
```

@@

Next, we will look at the gene set scores for three different `EWS-FLI1` gene sets. hierarchical
The below plots will all use the `mean-GENESET` columns.

@@

```{r}
# export final TSV with annotations 
final_classifications <- classification_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    tumor_cell_classification = combined_cnv_classification
  ) |> 
  unique()

>>>>

# output files 
final_annotations_dir <- file.path(module_base, "results", "annotation_tables", "SCPCS000490")
fs::dir_create(final_annotations_dir)
final_annotations_file <- file.path(final_annotations_dir, "SCPCL000822_tumor-classifications.tsv.gz")
```

@@

# generate marker genes df
plot_markers_df <- create_marker_gene_df(
  sce,
  classification_df, 
  marker_genes_file
)
```

@@

ggplot(marker_density_df, aes(x = sum_raw_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method))
```

@@

Next, we will look at the gene set scores for three different `EWS-FLI1` gene sets. heirarchical
The below plots will all use the `mean-GENESET` columns.

@@

```{r}
# export final TSV with annotations 
final_classifications <- classification_df |> 
  dplyr::select(
    "cell_barcode" = barcodes,
    combined_cnv_classification
  ) |> 
  unique()

### 2024-07-16 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
update output tables for 822 and 824

# reference files to use with SingleR
ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
ref_geneset_scores_file <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
```

@@

# filter out any ambiguous cells 
caret_df <- ref_classification_df |> 
  dplyr::filter(tumor_cell_classification != "Ambiguous") |> 
  # make sure positive class is tumor
  dplyr::mutate(
    tumor_cell_classification = forcats::fct_relevel(tumor_cell_classification, "Tumor"),
    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
  )

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      caret_df$tumor_cell_classification, 
      caret_df$auc_classification)
  ) 
```

@@

# add classification column for comparison to original classification
caret_df <- caret_df |> 
  dplyr::mutate(
    ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal"), 
    tumor_cell_classification = forcats::fct_relevel(tumor_cell_classification, "Tumor"),
    ucell_classification = forcats::fct_relevel(ucell_classification, "Tumor")
  )

@@

Finally, we will look at the marker gene and gene set scores for every cell across annotation methods.

```{r}
# create annotation df, keeping all classification methods 
annotation_df <- classification_df |> 
  dplyr::select(barcodes, marker_gene_classification, updated_marker_gene_classification, auc_classification, ucell_classification) |> 
  unique()

@@

annotation <- ComplexHeatmap::columnAnnotation(
  marker_genes = annotation_df$marker_gene_classification,
  updated_marker_genes = annotation_df$updated_marker_gene_classification,
  AUCell = annotation_df$auc_classification, 
  UCell = annotation_df$ucell_classification,
  col = list(marker_genes = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
             updated_marker_genes = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"), 
             AUCell = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
             UCell = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
```

@@

```{r}
# export final TSV with annotations 
classifications_output <- classification_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    marker_gene_classification,
    updated_marker_gene_classification, 
    auc_classification,
    ucell_classification
  ) |> 
  unique()

>>>>

# reference files to use with SingleR
ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv.gz")
ref_geneset_scores_file <- file.path(module_base, "results", "cnv_annotation", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
```

@@

# filter out any ambiguous cells 
caret_df <- ref_classification_df |> 
  dplyr::filter(combined_cnv_classification != "Ambiguous") |> 
  # make sure positive class is tumor
  dplyr::mutate(
    cnv_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor"),
    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
  )

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      caret_df$cnv_classification, 
      caret_df$auc_classification)
  ) 
```

@@

# add classification column for comparison to original classification
caret_df <- caret_df |> 
  dplyr::mutate(
    ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal"), 
    tumor_cell_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor"),
    ucell_classification = forcats::fct_relevel(ucell_classification, "Tumor")
  )

@@

Finally, we will look at the marker gene and gene set scores for every cell across annotation methods. 
We will also add in annotations that are the consensus between the updated marker gene and `AUCell`. 
Cells that are called tumor in both will be "Tumor" and normal in both will be called "Normal". 
All other cells will be labeled as "Ambiguous".

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(consensus_classification = dplyr::case_when(
    updated_marker_gene_classification == "Tumor" & auc_classification == "Tumor" ~ "Tumor",
    updated_marker_gene_classification == "Normal" & auc_classification == "Normal" ~ "Normal",
    .default = "Ambiguous"
  ))
```

```{r}
# create annotation df, keeping all classification methods 
annotation_df <- classification_df |> 
  dplyr::select(barcodes, marker_gene_classification, updated_marker_gene_classification, auc_classification, ucell_classification, consensus_classification) |> 
  unique()

@@

plot_colors <- c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")

annotation <- ComplexHeatmap::columnAnnotation(
  marker_genes = annotation_df$marker_gene_classification,
  updated_marker_genes = annotation_df$updated_marker_gene_classification,
  AUCell = annotation_df$auc_classification, 
  UCell = annotation_df$ucell_classification,
  consensus = annotation_df$consensus_classification,
  col = list(marker_genes = plot_colors,
             updated_marker_genes =plot_colors, 
             AUCell = plot_colors,
             UCell = plot_colors,
             consensus = plot_colors))
```

@@

```{r}
# export final TSV with annotations 
classifications_output <- classification_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    marker_gene_classification,
    updated_marker_gene_classification, 
    auc_classification,
    ucell_classification,
    consensus_classification
  ) |> 
  unique()
